---
title: "Gapminder"
author: "Anmol Rattan Singh Sandhu"
date: 2023-02-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
gapminder %>% glimpse()
```

**Observations**:

-   The dataset has the following variables:
    -   country

    -   continent

    -   year

    -   lifeExp

    -   pop

    -   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% 
  pull(year) %>% 
  max()
year_min <- gapminder %>% 
  pull(year) %>% 
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

**Observations**:

-   The GDP has quite a few outliers and is skewed towards higher values for every continent. This is especially noticeable for Asia.

**Difficulties & Approaches**:

-   I tried using a scatter plot at first but the data was too dense near the bottom and it was hard to make sense of it.
-   I noticed a few outliers in the scatter plot and decided to use a box plot to visualize them.
-   This was a good approach but it still was not the best since the outliers made it difficult to analyze the lower GDP values in a continent.
-   The final thing I did to make the plot more readable was adding a log scale on the y-axis.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
gapminder %>% 
  filter(year == year_min) %>% 
  arrange(desc(gdpPercap)) %>% 
  group_by(continent) %>% 
  slice(1:4)
```

**Observations**:

-   Identify the outlier countries from q2
    -   After sorting the countries by `gdpPercap` we can look at the top countries grouped by continent to find the outliers.
    -   Some of the outliers are South Africa, Gabon, United States, Canada, Kuwait, Bahrain, etc.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
gapminder %>%
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("South Africa", "Gabon", "United States",
                                       "Canada", "Kuwait", "Bahrain")),
    mapping = aes(color = country),
    size = 2
  ) +
  scale_y_log10() +
  facet_wrap(~year)
```

**Observations**:

-   Most of the top outlier countries in 1952 are still the top countries in 2007.
-   The `gdpPercap` is more spread out in 2007 compared to 1952 which implies wealth difference between countries in higher in 2007.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder %>%
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

-   `lifeExp` varies by a significant amount for countries in different continents but the variation range is similar for most continents except Asia. Africa and Europe are the only ones with outliers but interestingly Europe has lower outliers than Africa does.

    ```{r}
    top_lifeExp <- gapminder %>% 
      filter(year == year_max) %>% 
      arrange(desc(lifeExp)) %>% 
      group_by(continent) %>% 
      slice(1:2) %>% 
      pull(country)

    best_lifeExp <- gapminder %>% 
      filter(country %in% top_lifeExp)
    best_lifeExp
    ```

```{r q5-task2}
best_lifeExp %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point(aes(color = country))
```

-   The average `lifeExp` consistently went up for all the top countries from 1952 to 2007.
-   The `lifeExp` had a lot of variation between different countries in 1950 but the variation is much lower in 2007.
-   Countries with a lower `lifeExp` experienced a much higher increase compared to countries with an already high `lifeExp`.

```{r q5-task3}
best_lifeExp %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point(aes(color = country)) + 
  geom_point(
    data = . %>% filter(year == year_min),
    aes(size = gdpPercap)
    )
```

-   Countries with a higher `gdpPercap` had a higher `lifeExp` in 1952 compared to countries with a lower `gdpPercap` which shows a direct correlation between `lifeExp` and `gdpPercap` in 1952.
