---
title: "SAT and College Grades"
author: "Anmol Rattan Singh Sandhu"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: How do we apply hypothesis testing to investigating data? In
this challenge you'll practice using hypothesis testing tools to make
sense of a dataset.

*Reading*: - [Harvard Study Says SATs Should Be Optional: Here's
Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why)
(Optional); easy-to-read news article on colleges going SAT-free -
[Norm-Referenced Tests and Race-Blind
Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university)
(Optional); technical report on relationship between the SAT/ACT and
non-academic factors

*Credit*: This is based on a [case
study](http://onlinestatbook.com/2/case_studies/sat.html) originally
prepared by Emily Zitek, with data collected through the research of
Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
```

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

*Background*: Every year about 2 million students take the Scholastic
Aptitude Test (SAT). The exam is
[controversial](http://www.nea.org/home/73288.htm) but [extremely
consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html).
There are many claims about the SAT, but we're going to look at just
one: Is the SAT predictive of scholastic performance in college? It
turns out this is a fairly complicated question to assess---we'll get an
introduction to some of the complexities.

# Obtain the Data

<!-- -------------------------------------------------- -->

### **q1** Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_excel("data/sat.xls")

## TODO: Do your "first checks"
df_sat %>% glimpse()
```

**Observations**:

-   Fill in the following "data dictionary"

| Column     | Meaning                         |
|------------|---------------------------------|
| `high_GPA` | The student's high school GPA   |
| `math_SAT` | Score in the SAT math section   |
| `verb_SAT` | Score in the SAT verbal section |
| `comp_GPA` | The student's CS GPA            |
| `univ_GPA` | The student's university GPA    |

-   What information do we have about these students?
    -   The data contains information like high school GPAs, SAT scores,
        and college grades.
-   What kinds of information *do we not have* about these students?
    -   We do not have any other demographic information about the
        students other than their scores and grades.
-   Based on these missing variables, what possible effects could be
    present in the data that we would have *no way of detecting*?
    -   Information like access to resources, which high school the
        student went to, family income, etc. allows us to only perform a
        narrow data analysis.

# Analysis with Hypothesis Testing

<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the
first based on hypothesis testing of correlation coefficients, and the
second based on fitting a regression model and interpreting the
regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations

<!-- ----------------------------------------------------------------------- -->

### **q2** Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
# create a scatter plot with univ_GPA on the y-axis, and high_GPA and both_SAT on the x-axis
df_composite %>% 
  ggplot(aes(x = univ_GPA)) +
    geom_point(aes(y = high_GPA, color = "high_GPA")) +
    geom_point(aes(y = both_SAT * (4.0/1600), color = "both_SAT")) +
    xlab("University GPA") +
    ylab("High School GPA") +
    scale_color_manual(name = "Variables", values = c("blue", "red")) +
    ggtitle("Comparison of Univ_GPA against High_GPA and Both_SAT") +
    # Add a second y-axis for both_SAT
    scale_y_continuous(
      name = "High School GPA",
      sec.axis = sec_axis(~.*(1600/4.0), name = "SAT Scores")
    )
```

**Observations**:

-   What relationship do `univ_GPA` and `both_SAT` exhibit?
    -   They show a positive correlation.
-   What relationship do `univ_GPA` and `high_GPA` exhibit?
    -   They also show a positive correlation but it is a stronger
        correlation compared to `both_SAT`.

### Hypothesis Testing with a Correlation Coefficient

<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation
coefficient. The idea is to set our null hypothesis to the case where
there is no correlation, and test to see if the data contradict that
perspective. Formally, the null (H0) and alternative (HA) hypotheses
relating to a correlation coefficient between two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the
assumption that `X, Y` are both normally distributed. First, let's check
to see if this assumption looks reasonable for our data.

### **q3** Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
# pivot the data
df_composite %>% 
  pivot_longer(
    cols = c(high_GPA, both_SAT, univ_GPA), 
    names_to = "Variable", 
    values_to = "Value"
  ) %>% 
  ggplot(aes(x = Value)) +
  geom_histogram(alpha = 0.5, bins = 15) +
  facet_grid(col = vars(Variable), scales = "free") +
  ggtitle("Histograms of Both_SAT, High_GPA, and Univ_GPA") +
  labs(x = "Value", y = "Frequency")

```

**Observations**:

-   To what extent does `both_SAT` look like a normal distribution?
    -   `both_SAT` does not completely look like a normal distribution.
        It is slightly right-skewed.
-   To what extent does `high_GPA` look like a normal distribution?
    -   `high_GPA` is the closet one to a normal distribution but it is
        also slightly left-skewed.
-   To what extent does `univ_GPA` look like a normal distribution?
    -   `univ_GPA` does not look like a normal distribution, it is very
        left-skewed.

Keep in mind your findings as you complete q4.

### **q4** Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
cor.test(pull(df_composite, high_GPA), pull(df_composite, univ_GPA))
cor.test(pull(df_composite, both_SAT), pull(df_composite, univ_GPA))

```

**Observations**:

-   Which correlations are significantly nonzero?
    -   Both of the correlations are significantly nonzero at 0.7795631
        and 0.6846776.
-   Which of `high_GPA` and `both_SAT` seems to be more strongly
    correlated with `univ_GPA`?
    -   `high_GPA` has a stronger correlation with `univ_GPA`.
-   How do the results here compare with the visual you created in q2?
    -   It's hard to see in the scatter plot but the `high_GPA` does
        seem to increase slightly more strongly with the `univ_GPA`.
-   Based on these results, what can we say about the predictive
    capabilities of both `high_GPA` and `both_SAT` to predict
    `univ_GPA`?
    -   `high_GPA` does have a stronger correlation than `both_SAT` so
        it's the better one between the two. Both of them do have a
        significant correlation with `univ_GPA` so they do offer some
        predictive value.

Finally, let's use the bootstrap to perform the same test using
*different* assumptions.

### **q5** Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
high_GPA <-
  df_composite %>% 
  pull(high_GPA)

univ_GPA <-
  df_composite %>% 
  pull(univ_GPA)

both_SAT <-
  df_composite %>% 
  pull(both_SAT)

n_boot <- 1000

bootstrap_corr <- numeric(n_boot)
for (i in 1:n_boot) {
  sample_idx <- sample(length(high_GPA), replace = TRUE)
  sample_high_GPA <- high_GPA[sample_idx]
  sample_univ_GPA <- univ_GPA[sample_idx]

  bootstrap_corr[i] <- cor(sample_high_GPA, sample_univ_GPA)
}

# Compute confidence interval
lower_ci <- quantile(bootstrap_corr, 0.025)
upper_ci <- quantile(bootstrap_corr, 0.975)
est <- quantile(bootstrap_corr, 0.5)

tibble(
  lower_ci = lower_ci,
  estimate = est,
  upper_ci = upper_ci)

```

**Observations**:

-   How does your estimate from q5 compare with your estimate from q4?
    -   The estimate in q5 is close to the one in q4.
-   How does your CI from q5 compare with your CI from q4?
    -   The upper bound is higher in q5 than q4.

*Aside*: When you use two different approximations to compute the same
quantity and get similar results, that's an *encouraging sign*. Such an
outcome lends a bit more credibility to the results.

## View 2: Modeling

<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study
the relationship among more variables we can instead use a fitted model.
Using a model, we can also help assess whether it is *worthwhile* to
measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model

<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a
model, we can express our hypotheses in terms of the model parameters.
For instance, if we were interested in whether $X$ has an affect on $Y$,
we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on
$Y$. Let's apply this idea to relating the variables `univ_GPA` and
`high_GPA`. Luckily R has built-in tools to construct a confidence
interval on the $\beta$'s in a regression [1]; we'll simply use those
tools rather than do it by hand.

### **q6** Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <-
  df_composite %>% 
  lm(formula = univ_GPA ~ both_SAT)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

-   What is the confidence interval on the coefficient of `both_SAT`? Is
    this coefficient significantly different from zero?
    -   The confidence interval is approximately between 0.0020 and
        0.0035. This is significantly different from zero since the
        interval is above zero.
-   By itself, how well does `both_SAT` predict `univ_GPA`?
    -   It has a noticeable positive correlation therefore it does offer
        some predictive value. `high_GPA` has a higher correlation with
        `univ_GPA` and might offer even better for predictive value.

Remember from `e-model03-interp-warnings` that there are challenges with
interpreting regression coefficients! Let's investigate that idea
further.

### **q7** Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
df_composite %>% 
  lm(formula = univ_GPA ~ high_GPA + both_SAT) %>% 
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

df_composite %>% 
  lm(formula = univ_GPA ~ high_GPA) %>% 
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

-   How well do these models perform, compared to the one you built in
    q6?
    -   They perform better, outperforming the one in q6.
-   What is the confidence interval on the coefficient of `both_SAT`
    when including `high_GPA` as a predictor?? Is this coefficient
    significantly different from zero?
    -   This interval does include zero therefore the coefficient is not
        significantly different from zero.
-   How do the hypothesis test results compare with the results in q6?
    -   This model does perform better but also have zero in it's
        confidence interval making it less reliable.

## Synthesize

<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### **q8** Using the results from all previous q's, answer the following questions.

**Observations**:

-   Between `both_SAT` and `high_GPA`, which single variable would you
    choose to predict `univ_GPA`? Why?
    -   If we had to choose only one variable to predict **`univ_GPA`**
        between **`high_GPA`** and **`both_SAT`**, I would choose
        **`high_GPA`**.
    -   Firstly, when we visually examined the scatterplot between
        **`high_GPA`** and **`univ_GPA`**, we observed a stronger linear
        relationship between the two variables compared to
        **`both_SAT`** and **`univ_GPA`**.
    -   Secondly, when we fit a linear regression model with each
        variable separately and compared the accuracy of predictions
        using validation data, the model with **`high_GPA`** had a lower
        error than the model with **`both_SAT`**.
    -   It's worth noting that correlation does not necessarily imply
        causation, and there may be other factors that are important in
        predicting **`univ_GPA`** which are not included in the data.
-   Is `both_SAT` an effective predictor of `univ_GPA`? What specific
    pieces of evidence do you have in favor of `both_SAT` being
    effective? What specific pieces of evidence do you have against?
    -   `both_SAT` does have a positive correlation with `univ_GPA` but
        the analysis suggests that `high_GPA` is a better predictor of
        `univ_GPA`. The main evidence against that is the higher
        positive correlation `high_GPA` has with `univ_GPA`.

# End Notes

<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more
information see this [Wiki
article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
