---
title: "Massachusetts Highway Stops"
author: "Anmol Rattan Singh Sandhu"
date: 2023-05-03
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r}
summary(df_data)
```

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
df_data %>% 
  pull(subject_race) %>% 
  unique()

df_data %>% 
  pull(raw_Race) %>% 
  unique()
```

**Observations**:

-   What are the unique values for `subject_race`?
    -   White, Hispanic, Black, Asian/Pacific Islander, other, NA and unknown
-   What are the unique values for `raw_Race`?
    -   White, Hispanic, Black, Asian or Pacific Islander, Middle Eastern or East Indian(South Asian), American Indian or Alaskan Native, NA, None - for no operator present citations only, A
-   What is the overlap between the two sets?
    -   White, Hispanic, Black, Asian or Pacific islander, NA
-   What is the difference between the two sets?
    -   Middle Eastern or East Indian(South Asian), American Indian or Alaskan Native, None - for no operator present citations only, A, other, unknown

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
df_data %>%
  mutate(
    raw_Race = tolower(raw_Race),
    subject_race = as.character(subject_race),
    diff = +(raw_Race != subject_race),
    same = +(raw_Race == subject_race)
  ) %>%
  summarise(
    num_diff = sum(diff, na.rm = TRUE),
    num_same = sum(same, na.rm = TRUE)
  ) %>%
  mutate(
    percentage = num_same/(num_diff + num_same)
  ) %>%
  glimpse()

```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   It is the most plausible that race_Raw is the unprocessed version of subject_race since they have a 95% match. It is possible that 5% of the values were updated in processing.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r}
df_data %>%
  filter(!is.na(subject_age)) %>%
  group_by(subject_age) %>%
  summarise(arrest_rate = mean(arrest_made, na.rm = TRUE)) %>%
  ggplot(aes(x = subject_age, y = arrest_rate)) +
  geom_line() +
  labs(x = "Subject age", y = "Arrest rate")

```

```{r}
df_data %>%
  filter(!is.na(subject_sex)) %>%
  group_by(subject_sex) %>%
  summarise(arrest_rate = mean(arrest_made, na.rm = TRUE)) %>%
  ggplot(aes(x = subject_sex, y = arrest_rate, fill = subject_sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Subject sex", y = "Arrest rate", fill = "Subject sex")

```

```{r}
df_data %>%
  filter(!is.na(subject_race)) %>%
  group_by(subject_race) %>%
  summarise(arrest_rate = mean(arrest_made, na.rm = TRUE)) %>%
  ggplot(aes(x = subject_race, y = arrest_rate, fill = subject_race)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Subject race", y = "Arrest rate", fill = "Subject race")
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?
    -   The age range is very right skewed peaking in the teenager range with a sudden decrease. Then it slowly rises again and slowly decreased with age.
-   How does `arrest_rate` tend to vary with `subject_sex`?
    -   The arrest rate of male individuals is twice of the arrest rate of female individuals.
-   How does `arrest_rate` tend to vary with `subject_race`?
    -   Hispanics have significantly higher arrest rate followed by black and white. The other category seems interesting since nothing is mentioned about it.

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   white, black and Hispanic.
-   Which `subject_race` levels have terms in the model?
    -   Hispanic and white.

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
fit_q7 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex,
  data = df_data %>%
    filter(
      !is.na(arrest_made),
      subject_race %in% c("white", "black", "hispanic")
    ) %>% 
    mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

fit_q7 %>% 
  tidy()

fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   The highest rate is if the `subject_race` is Hispanic and the lowest is white since that is set as the reference.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   Racial Bias of police officer.
    -   Population diversity differences.
    -   Economic differences.
-   Look at the sent of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   A few columns might be interesting to look at such as `Reason for stop`, `county_name`, `search_basis`, `contraband_found`, etc.

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
  data = df_data %>%
    filter(
      !is.na(arrest_made), 
      subject_race %in% c("white", "black", "hispanic")) %>%
    mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   The data shows that Hispanic subjects are more likely to be arrested, followed by black subjects, and then white subjects.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   While the presence of contraband is a stronger predictor of arrest rate compared to race, there is still evidence of a significant association between arrest rate and race for Hispanic individuals even after controlling for contraband presence.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

How do different counties vary by arrest rates?

```{r}
fit_q9 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found + county_name,
  data = df_data %>%
    filter(
      !is.na(arrest_made), 
      subject_race %in% c("white", "black", "hispanic")) %>%
    mutate(
      subject_race = fct_relevel(subject_race, "white"),
      county_name = fct_relevel(county_name, "Worcester County")
      ),
  family = "binomial"
)

fit_q9 %>% tidy()

```

**Observations**:

-   Highest arrest rate:
    -   I picked one of the counties as the baseline.
    -   Berkshire County and Essex County had the highest arrest rate.
    -   Dukes County and Hampshire County had the lowest arrest rate.
-   I have other questions like are there any variables in the model that moderate the relationship between county and arrest rate? For example, do certain races or age groups have significantly different arrest rates in different counties?
    -   I tried factoring all those in the model but that returned a lot of row ratios which were hard to parse. I think I need to get more practice with modelling.

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
